Shakespeare Corpus Analysis

```{r}
#Load required libraries
library(dplyr)
library(ggplot2)
library(languageR)
library(slam)
library(tm)
```

```{r}
#Load and summarize data
data <- read.csv(file.choose(), header = TRUE)
summary(data)
```

a. Comparative Analysis of Familiar ('Thou/Thee') and Formal ('You/Ye(e)') Pronoun Usage":  Which plays included more instances of the familiar 2p pronoun ("thou/thee")
relative to the formal 2p pronoun "you/ye(e)"?

```{r}
#Filtering the "yee" variant
grep("\\<(yee)\\>", tolower(data$PlayerLine), value = TRUE)
```
Apparently, the variant "yee" does not occur in any of the plays so it will not be 
considered here any further.


```{r}
# Create vector containing unique values from Play column
plays <- unique(data$Play) 
# Create vectors to store the frequencies of "thou", "thee", "you" and "ye"
thou_freq<-numeric()
thee_freq<-numeric()
you_freq<-numeric()
ye_freq<-numeric()

# Loop over each unique play
for (i in seq_along(plays)){ 

#Extraction and combination of PlayerLine into a single text
  text <- Corpus(VectorSource(paste(data[data$Play==plays[i],]$PlayerLine,collapse=" ")))
# Remove punctuation
  text <- tm_map(text, removePunctuation) 
#Create a term-document matrix (tdm), treating words as terms
  tdm  <- TermDocumentMatrix(text, control=list(stopwords=FALSE, wordLengths=c(2, Inf)))

#Counting occurrences of each pronoun and storing them in respective vector
  thou_freq[i]<-as.numeric(slam::row_sums(tdm)["thou"])
  thee_freq[i]<-as.numeric(slam::row_sums(tdm)["thee"])
  you_freq[i]<-as.numeric(slam::row_sums(tdm)["you"])
  ye_freq[i]<-as.numeric(slam::row_sums(tdm)["ye"])
}
```


```{r}
# Create dataframe containing plays and frequencies of pronouns
pronoun_freq <- data.frame(plays, thou_freq, thee_freq, you_freq, ye_freq) 
# Remove missing values
pronoun_freq <- na.omit(pronoun_freq) 
# Sort dataframe by frequency in descending order
pronoun_freq <-pronoun_freq[order(-pronoun_freq$thou_freq, 
                                  -pronoun_freq$thee_freq, 
                                  -pronoun_freq$you_freq, 
                                  -pronoun_freq$ye_freq),] 
pronoun_freq
```

#Create summary table of pronoun usage, adding informal and formal counts, filtering for plays where informal pronouns are more prevalent than formal pronouns
```{r}
pronoun_table <-
  pronoun_freq %>%
  mutate(informal = thou_freq + thee_freq) %>%
  mutate(formal = you_freq + ye_freq) %>%
  group_by(plays, informal, formal) %>%
  filter(informal > formal) %>%
  summarise()
pronoun_table

```
The data frame pronoun_table shows the 10 plays which include more instances of the familiar 2P pronoun relative to the formal 2P pronoun. This might be due to socio-linguistic reasons: the majority of these plays feature protagonists who are at the top of society (usually a king, Timon of Athens was a member of the high society) so there is little reason for these protagonists to address anybody in the formal way. Most of the people they address are either in a socially equal or inferior position. However, other factors such as play genre, the speaker’s emotional state, or rhetorical choices may also play a role in pronoun usage and could be explored further.


b. Analysis of the Use of Familiar 2p Pronoun and Formal 2p Pronoun Across Characters 
```{r}
# Create vector containing unique values from Player column
player <- unique(data$Player) 
# Create vectors to store the frequencies of "thou", "thee", "you" and "ye" per player
thou_freq_player<-numeric()
thee_freq_player<-numeric()
you_freq_player<-numeric()
ye_freq_player<-numeric()

```

```{r}
## Loop over each unique player
for (i in seq_along(player)){ 
  # Combine all lines spoken by the player into text corpus
  text <- Corpus(VectorSource(paste(data[data$Player==player[i],]$PlayerLine,collapse=" ")))
  # Remove punctuation
  text <- tm_map(text, removePunctuation) 
  #Create a term-document matrix (tdm), treating words as terms
  tdm  <- TermDocumentMatrix(text, control=list(stopwords=FALSE, wordLengths=c(2, Inf)))
   # Calculate frequencies for each pronoun
  thou_freq_player[i]<-as.numeric(slam::row_sums(tdm)["thou"]) 
  thee_freq_player[i]<-as.numeric(slam::row_sums(tdm)["thee"]) 
  you_freq_player[i]<-as.numeric(slam::row_sums(tdm)["you"]) 
  ye_freq_player[i]<-as.numeric(slam::row_sums(tdm)["ye"]) 
}
```

```{r}
# Create dataframe containing player names and frequencies of pronouns
pronoun_freq_player <- data.frame(player, thou_freq_player, thee_freq_player, you_freq_player, ye_freq_player) 
# Remove missing values
pronoun_freq_player <- na.omit(pronoun_freq_player) 
# Sort dataframe by frequency in descending order
pronoun_freq_player <-pronoun_freq_player[order(-pronoun_freq_player$thou_freq_player
                                                -pronoun_freq_player$thee_freq_player,
                                                -pronoun_freq_player$you_freq_player, 
                                                -pronoun_freq_player$ye_freq_player),]
pronoun_freq_player 
```

```{r}
# Filter players who predominantly use informal pronouns and sort by frequency
informal_freq_player <- pronoun_freq_player %>%
  mutate(informal_player = thou_freq_player + thee_freq_player) %>%
  mutate(formal_player = you_freq_player + ye_freq_player) %>%
  group_by(player, informal_player, formal_player) %>%
  filter(informal_player > formal_player) %>%
  summarise()
informal_freq_player

# Sort the data frame by informal pronoun frequency in descending order
informal_freq_player <-informal_freq_player[order(-informal_freq_player$informal_player),] 
informal_freq_player

```
The data frame informal_freq_player shows the 32 characters who use the informal variant of the 2P pronoun  more frequently than the formal Among the first 10 most frequent hits, the majority is, once again, regents (or in Timon's case a member of the high society). 
An exception to this trend is Falstaff, but considering the fact that Falstaff is a close companion to the king, it is not surprising that he uses the informal pronoun so often as their relationship is a personal one.

```{r}
# Filter players who predominantly use formal pronouns and sort by frequency
formal_freq_player <- pronoun_freq_player %>%
  mutate(informal_player = thou_freq_player + thee_freq_player) %>%
  mutate(formal_player = you_freq_player + ye_freq_player) %>%
  group_by(player, informal_player, formal_player) %>%
  filter(formal_player > informal_player) %>%
  summarise()
formal_freq_player
# Sort the resulting data frame by formal pronoun frequency in descending order
  formal_freq_player <-formal_freq_player[order(-formal_freq_player$formal_player),] 
  formal_freq_player
```
The dataframe formal_freq_player lists the 66 characters who use the formal 2P pronoun more often. Among the first 10 most frequent hits are some regents, in these cases a possible explanation for the use of the formal pronoun are speaking in public addressing a group of people. This might  also be the case for Brutus. The formal pronoun is also employed to address people in  a higher social position and this fact might explain Hamlet's and Helena's use of you/ye. 
